---
title: "hw04-tidy data and joins"
author: "Frederike Basedow"
date: "7 Oktober 2018"
output: github_document
---


# Homework 04: Tidy data and joins

## Overview

Due Tuesday 2018-10-09 at 23:59.

You have learned alot about data wrangling! You know how to use the tidyverse to:

* Filter rows and select variables in a principled or name-based manner
* Modify variables and create new variables
* Group rows of a data frame and do summarizing or mutating computations within group
* Reshape a data frame
* Get new information into a data frame via a joining

The goal of this homework is to solidify your data wrangling skills by working some realistic problems in the grey area between data aggregation and data reshaping.

If you internalize that there are multiple solutions to most problems, you will spend less time banging your head against the wall in data analysis. If something's really hard, sneak up on it from a different angle.

### Evaluation

All rubrics listed on the [assignments](http://stat545.com/Classroom/assignments/) page are relevant for this assignment.

## The Assignment

This is a "choose your own adventure"-style assignment, where you are expected to do the following two things:

1. Pick one of the data reshaping prompts and do it.
2. Pick one of the join prompts and do it.

It is fine to work with a new dataset and/or create variations on these problem themes.

```{r, message=FALSE}
library(gapminder)
library(tidyverse)
library(knitr)
```


### Data Reshaping Prompts (and relationship to aggregation)

__Problem__: You have data in one "shape" but you wish it were in another. Usually this is because the alternative shape is superior for presenting a table, making a figure, or doing aggregation and statistical analysis.

__Solution__: Reshape your data. For simple reshaping, `gather()` and `spread()` from `tidyr` will suffice. Do the thing that is possible / easier now that your data has a new shape.

__Prompts__:

Activity #1

* Make your own cheatsheet similar to Tyler Rinker's [minimal guide to `tidyr`](https://github.com/trinker/tidyr_in_a_nutshell).

Activity #2

* Make a tibble with one row per year and columns for life expectancy for two or more countries.
    - Use `knitr::kable()` to make this table look pretty in your rendered homework.
    - Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

I will first filter out the data from Germany and Canada in the `gapminder` data set.

```{r}
LE_C_Ge <- gapminder %>%
  group_by(year) %>%
  filter(country %in% c("Canada", "Germany")) %>%
  select(country, lifeExp, year) # select only important columns

kable(head(LE_C_Ge))
```


Next, I will `spread` this data so that it is easier to view in a table. I will make one column per country, which will contain the life expectancy data for that country:

```{r}
untidy_LE_C_Ge <- spread(LE_C_Ge, key = "country", value = lifeExp)

kable(untidy_LE_C_Ge, col.names = c("Year", "Canada", "Germany"))
```

This table shows the life expectancy in Canada in Germany per year. I will plot this in a scatter plot to see if there is a correlation in life expectany between Canada and Germany. I will play around with the `theme` options from `ggplot2` for practice.

```{r}
untidy_LE_C_Ge %>%
  ggplot(aes(Germany, Canada)) +
  geom_point(size = 2, colour = "orange") +
  geom_smooth(method = lm, se = FALSE, size = 0.7, colour = "hotpink") +
  labs(title = "Correlation of Life Expectancy in Germany vs. Canada") +
  theme(panel.background = element_rect(fill = "white", 
                                        colour = "grey40", 
                                        size = 1),
        panel.grid.major = element_line(colour = "grey80", 
                                        size = 0.3), 
        axis.text = element_text(colour = "grey40", 
                                 size = 10), 
        axis.ticks = element_line(size = 2, 
                                  colour = "grey40"), 
        axis.title.y = element_text(size = 12, 
                                    colour = "grey40",
                                    face = "bold"), 
        axis.title.x = element_text(size = 12, 
                                    colour = "grey40",
                                    face = "bold"), 
        plot.title = element_text(size = 15, 
                                  colour = "grey20",
                                  face = "bold"))
```

Looks like there's a positive correlation in life expectancy between Canada and Germany.


Activity #3

* Compute some measure of life expectancy (mean? median? min? max?) for all possible combinations of continent and year. Reshape that to have one row per year and one variable for each continent. Or the other way around: one row per continent and one variable per year.
    - Use `knitr::kable()` to make these tables look pretty in your rendered homework.
    - Is there a plot that is easier to make with the data in this shape versis the usual form? If so (or you think so), try it! Reflect.

Activity #4

* In [Window functions](http://stat545.com/block010_dplyr-end-single-table.html#window-functions), we formed a tibble with 24 rows: 2 per year, giving the country with both the lowest and highest life expectancy (in Asia). Take that table (or a similar one for all continents) and reshape it so you have one row per year or per year * continent combination.

Activity #5

* Previous TA Andrew MacDonald has a nice [data manipulation sampler](https://gist.github.com/aammd/11386424). Make up a similar set of exercises for yourself, in the abstract or (even better) using Gapminder or other data, and solve them.

### Join Prompts (join, merge, look up)

__Problem__: You have two data sources and you need info from both in one new data object.

__Solution__: Perform a __join__, which borrows terminology from the database world, specifically SQL.

__Prompts__:

Activity #1

* Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a `dplyr` join function and make some observations about the process and result. Explore the different types of joins. Examples of a second data frame you could build:
    - One row per country, a country variable and one or more variables with extra info, such as language spoken, NATO membership, national animal, or capitol city.
    - One row per continent, a continent variable and one or more variables with extra info, such as northern versus southern hemisphere.

I had a look at the data sets included in R, using the `data() function` and found the `WorldPhones` data set, which includes the number of phones in each continent in different years. 

It may be interesting to see if the number of phones is related to the population in each continent. To have a look at this, I will add the `WorldPhones` information to a subset of `gapminder` that includes the mean population in each continent per year.

Let's have a look at the `WorldPhones` data set first:

```{r}
kable(WorldPhones)

class(WorldPhones)
```

First, I will have to get `WorldPhones` ready for its merge with `gapminder`. It is a matrix, which is hard to work with, so I will change it into a data frame. The years are currently rownames, so I will make them a column, as they are in `gapminder`. Finally, I will combine North America, South America and Middle America to one continent "Americas" as it is used in the gapminder data set.

```{r}
new_WP <- WorldPhones %>%
  as.data.frame() %>% 
  rownames_to_column("year") %>%
  mutate(Americas = N.Amer + S.Amer + Mid.Amer) %>% 
  select(-N.Amer, -S.Amer, -Mid.Amer) # deselect the orginal columns

kable(new_WP, col.names = c("Year", names(new_WP[2:6]))) # change column name of "year" column to "Year" for presentation with kable
```

Let's make it tidy so that it's the same style as `gapminder`. Instead of having one column per continent, we need one column that contains the different continents and one with the number of phones. We can use the `gather` function for that:

```{r}
tidy_newWP <- 
  new_WP %>% 
  gather(key="continent", value="n_phones", Europe:Americas) # if I don't specify which columns are part of the "key", the years will be taken as continents, too.

kable(head(tidy_newWP, 10))
```

Now we're almost ready to join, but there is still one problem. In order to join `tidy_newWP` with `gapminder`, the joining variables need to be the same type in the two data sets. I want to join by `year` and `continent` so let's check if these are the same in `gapminder` in my `tidy_newWP`.

```{r}
class(gapminder$year) == class(tidy_newWP$year)
class(gapminder$continent) == class(tidy_newWP$continent)
```

They are not the same in the two data sets. Let's see of what class they are:

```{r}
class(gapminder$year)
class(tidy_newWP$year)
class(gapminder$continent)
class(tidy_newWP$continent)
```

The `year` variable in `gapminder` is of the class integer and the `continent` variable is a factor, while both variables in `tidy_newWP` are characters. Let's adjust `tidy_newWP` so that the variable class matches the one in `gapminder`.

```{r}
tidy_newWP$year <- as.integer(tidy_newWP$year) 
tidy_newWP$continent <- as.factor(tidy_newWP$continent)

class(gapminder$year) == class(tidy_newWP$year)
class(gapminder$continent) == class(tidy_newWP$continent)

```

Great, now the `WorldPhones` data is ready to join! First, I will calculate the average population per continent per year in `gapminder`, which is what I want to combine with the `WorldPhones` data.

```{r}
pop_GM <-
  gapminder %>%
  group_by(year, continent) %>%
  summarize(mean_pop = mean(pop))

kable(head(pop_GM, 10))
```

Let's finally try to add the `WorldPhones` data to that! I will start with a `full_join` that combines all data from both data sets. I will join the data sets by both year and continent.

```{r}
pop_phone <- full_join(pop_GM, tidy_newWP, by = c("year", "continent"))

kable(head(pop_phone, 10)) # show first ten rows
kable(tail(pop_phone, 10)) # show last ten rows
```

Printing the ten first and last rows, we can see that there are a lot of missing values in both the `mean_pop` columns and the `n_phones` columnn. This is because `full_join` keeps all levels of both the `year` and `continent` variable, but only one of the years actually overlaps between the two data sets, as you can see here:

```{r}
intersect(gapminder$year, tidy_newWP$year)
```

Only year 1957 is present in both data sets. Let's join the data sets with only the data that is present in both. We can use the `inner_join` function for that.

```{r}
pop_phone_inner <- inner_join(pop_GM, tidy_newWP, by = c("year", "continent"))

kable(pop_phone_inner)
```

Now we have a short data frame that only contains the population and number of phones data from 1957 in each continent.

We can also figure out which data is the same in the two data sets by using the filtering join `semi_join`. 

```{r}
pop_phone_semi <- semi_join(pop_GM, tidy_newWP, by = c("year", "continent"))

kable(pop_phone_semi)
```

We can see that only data from year 1957 is kept, but the n_phones data from `tidy_newWP` is not added. If we flip around the order in which the data sets are entered we also only get data from 1957, but the gapminder-specific data is lacking:

```{r}
pop_phone_semi2 <- semi_join(tidy_newWP, pop_GM, by = c("year", "continent"))

kable(pop_phone_semi2)
```

We can also use `anti-join` to see which data is not overlapping, for example if we want to look at the number of phones in the years that are not included in gapminder:
```{r}
pop_phone_anti <- anti_join(tidy_newWP, pop_GM, by = c("year", "continent"))

kable(head(pop_phone_anti))
```

To try out all different joining options, I will use `left_join` and `right_join` next. `left_join` will only join the rows of `tidy_newWP` to `pop_GM` that are matching (if used in the order below).
```{r}
pop_phone_left <- left_join(pop_GM, tidy_newWP, by = c("year", "continent"))

kable(head(pop_phone_left, 20))
```

Similar to that is `right_join` which will only join rows of `pop_GM` that are matching in `tidy_newWP` (if used in the order below).

```{r}
pop_phone_right <- right_join(pop_GM, tidy_newWP, by = c("year", "continent"))

kable(head(pop_phone_right, 20))
```

To comparethe  population with number of phones, the `pop_phone_inner` output seems most useful. I will use it to plot mean population vs. number of phones. Note, that I could also use the output from `full_join`, `left_join`, or `right_join`.

```{r}
pop_phone_inner %>%  
  ggplot(aes(mean_pop, n_phones, colour = continent)) +
  geom_point(size = 3) +
  labs(title = "Number of Phones vs. Mean Population in 1957", 
       x = "Mean population", 
       y = "Number of Phones", 
       colour = "Continent:") +
  theme(legend.position = "bottom")
```

It looks like there is a higher number of phones in countries with higher population when looking at Africa, the Americas, Europe and Oceania, but Asia has a relatively low number of phones for its big population. However, we only have 5 independent values here and cannot draw any conclusions about a possible correlation. If we had numbers from different years we could see if the number of phones increases with an increasing population over time. 




Activity #2

* Create your own cheatsheet patterned [after Jenny's](http://stat545.com/bit001_dplyr-cheatsheet.html) but focused on something you care about more than comics! Inspirational examples:
    - Pets I have owned + breed + friendly vs. unfriendly + ??. Join to a table of pet breed, including variables for furry vs not furry, mammal true or false, etc.
    - Movies and studios....
    - Athletes and teams....

You will likely need to iterate between your data prep and your joining to make your explorations comprehensive and interesting. For example, you will want a specific amount (or lack) of overlap between the two data.frames, in order to demonstrate all the different joins. You will want both the data frames to be as small as possible, while still retaining the expository value.

Activity #3

* This is really an optional add-on to either of the previous activities.
* Explore the base R function `merge()`, which also does joins. Compare and contrast with `dplyr` joins.
* Explore the base R function `match()`, which is related to joins and merges, but is really more of a "table lookup". Compare and contrast with a true join/merge.

### Finishing up, and Reflection

Once you're done the above, go back to [UBC canvas](https://canvas.ubc.ca/), and find the "Homework 04" page.

You're encouraged to reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc. As usual, put this reflection on your canvas submission. It can be brief.

Please add a link to your homework respository to help out our wonderful TA's.